Flabby and injective objects in toposes

### 2018-10-21 paper-flabby-objects.tex
to do
Final draft

\begin{abstract}
  We introduce a general notion of \emph{flabby objects} in elementary toposes
  and study their basic properties. In the special case of localic toposes, this
  notion reduces to the common notion of flabby sheaves, yielding a
  site-independent characterization of flabby sheaves. Continuing a line of
  research started by Roswitha Harting, we use flabby objects to
  show that an internal notion of injective objects coincides with the
  corresponding external notion, in stark contrast with the situation for
  projective objects. We show as an application that higher direct images can
  be understood as internal cohomology, and study flabby objects in the
  \effective topos.
\end{abstract}

@@

{\centering \textbf{Rough draft, comments welcome! iblech@speicherleck.de}\par}
\bigskip

@@

The practicality of this approach hinges on the extent to which the dictionary
between internal and external notions has been worked out. For instance, the
simple example displayed above hinges on the dictionary entry stating that a
sheaf of modules is of finite type if and only if it looks like a
finitely-generated module from the internal point of view. The motivation for
this note was to find internal characterizations of flabby sheaves and of higher
direct images, and the resulting entries are laid out in
Section~\ref{sect:flabby-objects} and in
Section~\ref{sect:higher-direct-images}: A sheaf is flabby if and only if, from
the internal point of view, it is a flabby set, a notion introduced in
Section~\ref{sect:flabby-sets} below; and higher direct images look like sheaf
cohomology from the internal point view.

As a byproduct, we demonstrate how the notion of flabby sets is a useful
organizing principle in the study of injective objects. We employ flabby sets
to give a new proof of Roswitha Harting's result that injectivity of sheaves is
a local notion~\cite{harting:remark} and that a sheaf is injective if and only
if it is injective from the internal point of
view~\cite{harting:locally-injective}, which she stated (in slightly different
language) for sheaves of abelian groups. We use the opportunity to correct a
small mistake of her, namely claiming that no generalization to sheaves of
modules would be possible.

@@

In view of almost 80 years of sheaf cohomology, this state of affairs is
slightly embarassing, challenging the call that ``once [a] subject is better
understood, we can hope to refine its definitions and proofs so as to
avoid that axiom''~\cite[Section~3.4]{hott}.

A constructive account of sheaf cohomology would be highly desirable, not only
out of a philosophical desire to obtain a deeper understanding of the
foundations of sheaf cohomology, but also to: use the tools of sheaf cohomology
in the internal setting of toposes, thereby extending the applicability of
relativization by internalization; and to carry out \emph{integrated
developments} of algorithms for computing sheaf cohomology, where we would
extract algorithms together with termination and correctness proofs from a
hypothetical constructive account.

\textbf{Acknowledgments.} We are grateful to Thorsten Altenkirch, Simon Henry
and Maria Emilia Maietti for insightful discussions and pointers to prior work,
and to Jürgen Jost, Marc Nieper-Wißkirchen and Peter Schuster for valuable
guidance. Most of the work for this note was carried out at the University of
Augsburg and the Max Planck Institute for Mathematics in the Sciences in
Leipzig.

@@

There is an alternative definition of flabbiness, to be introduced below, which
is equivalent to the usual one in presence of Zorn's lemma and which
requires different commitments: For the alternative definition,
properties~(1),~(3b) and~(4a) can be verified purely intuitionistically, and
property~(2) can be verified purely intuitionistically for open injections~$f$.
There is a substitute for property~(3a) which can be verified purely
intuitionistically. We do not know whether property~(4b) can be established
purely intuitionistically, but we give a rudimentary analysis in
Section~\ref{sect:conclusion}.

@@

\begin{rem}Although we will see in Section~\ref{sect:in-eff} that there is some
relation between flabby sets and~$\neg\neg$-separated sets, neither notion
encompasses the other. The set~$\Omega$ is flabby, but might fail to
be~$\neg\neg$-separated; the set~$\NN$ is~$\neg\neg$-separated, even discrete,
but might fail to be flabby.
\end{rem}

\begin{defn}\begin{enumerate}
\item A set~$I$ is \emph{injective} if and only if, for any injection~$i
: A \to B$, any map~$f : A \to I$ can be extended to a map~$B \to I$.
\item An~$R$-module~$I$ is \emph{injective} if and only if, for any linear
injection~$i : A \to B$ between~$R$-modules, any linear map~$f : A \to I$ can
be extended to a linear map~$B \to I$.
\end{enumerate}
\[ \xymatrix{
  A \ar@{^{(}->}[r]\ar[d] & B \ar@{-->}@/^/[ld] \\
  I
} \]
\end{defn}

@@

\begin{proof}\begin{enumerate}
\item One can check that, for instance, the full powerset~$\P(X)$ and the set of
subterminals~$\P_{\leq 1}(X)$ are each injective.
\item The forgetful functor from modules to sets possesses a left exact left
adjoint. More explicitly, if~$i : A \to B$ is an injective map between sets and
if~$f : A \to I$ is an arbitrary map, then the induced map~$R\langle A \rangle
\to R\langle B \rangle$ between free modules is also injective, the given
map~$f$ lifts to a linear map~$R\langle A \rangle \to I$, and an~$R$-linear
extension~$R\langle B \rangle \to I$ induces an extension~$B \to M$ of~$f$.
\item One verifies that any abelian group embeds into a divisible
abelian group. By Baer's criterion (which requires the axiom of choice), divisible abelian groups are injective.
The result for modules over arbitrary rings then follows purely
formally~\cite{XXX}. \qedhere
\end{enumerate}\end{proof}

@@

Noticeably missing here is a statement as follows: ``Let~$0 \to M' \to M \to
M'' \to 0$ be a short exact sequence of modules. If~$M'$ and~$M$ are flabby, so
is~$M''$.'' Assuming Zorn's lemma in the metatheory, this statement is true in
every topos of sheaves over a locale, but we don't know whether it has an
intuitionistic proof and in fact we surmise that it has not.

@@

The assumption on~$f_*$ of Proposition~\ref{prop:pushforward-of-flabby-objects}
is for instance satisfied if~$f$ is a local geometric morphism or if~$f$ is
induced by an open continuous injection between topological spaces. XXX

@@

We originally set out to develop an intuitionistic account of Grothendieck's
sheaf cohomology. Čech methods can be carried out constructively, and
XXX(Barakat), but it appears that there is not a general framework for sheaf
cohomology which works in an intuitionistic metatheory.

@@

However, it is an open question under which circumstances quotient inductive
types can be shown to exist. Zermelo--Fraenkel with choice certainly suffices,
while Zermelo--Fraenkel without choice doesn't~\cite[Section~9]{shulman-lumsdaine:hits},
hence IZF also doesn't.\footnote{With quotient inductive types, any infinitary
algebraic theory admits free algebras. However, it is consistent with
Zermelo--Fraenkel set theory that some such theories don't admit free
algebras~\cite{blass:free-algebras}.} The existence of quotient inductive types
seem to be, as the existence of enough injective modules, \emph{constructively
neutral}.

@@

\printbibliography

>>>>

\begin{abstract}
  We introduce a general notion of \emph{flabby objects} in elementary toposes
  and study their basic properties. In the special case of localic toposes, this
  notion reduces to the common notion of flabby sheaves, yielding a
  site-independent characterization of flabby sheaves. Continuing a line of
  research started by Roswitha Harting, we use flabby objects to
  show that an internal notion of injective objects coincides with the
  corresponding external notion, in stark contrast with the situation for
  projective objects. We show as an application that higher direct images can
  be understood as internal cohomology, and we study flabby objects in the
  \effective topos.
\end{abstract}

@@

@@

The practicality of this approach hinges on the extent to which the dictionary
between internal and external notions has been worked out. For instance, the
simple example displayed above hinges on the dictionary entry stating that a
sheaf of modules is of finite type if and only if it looks like a
finitely-generated module from the internal point of view. The motivation for
this note was to find internal characterizations of flabby sheaves and of higher
direct images, and the resulting entries are laid out in
Section~\ref{sect:flabby-objects} and in
Section~\ref{sect:higher-direct-images}: A sheaf is flabby if and only if, from
the internal point of view, it is a flabby set, a notion introduced in
Section~\ref{sect:flabby-sets} below; and higher direct images look like sheaf
cohomology from the internal point of view.

As a byproduct, we demonstrate how the notion of flabby sets is a useful
organizing principle in the study of injective objects. We employ flabby sets
to give a new proof of Roswitha Harting's result that injectivity of sheaves is
a local notion~\cite{harting:remark} and that a sheaf is injective if and only
if it is injective from the internal point of
view~\cite{harting:locally-injective}, which she stated (in slightly different
language) for sheaves of abelian groups. We use the opportunity to correct a
small mistake of hers, namely claiming that no generalization to sheaves of
modules would be possible.

@@

In view of almost 80 years of sheaf cohomology, this state of affairs is
slightly embarassing, challenging the call that ``once [a] subject is better
understood, we can hope to refine its definitions and proofs so as to
avoid [the law of excluded middle]''~\cite[Section~3.4]{hott}.

A constructive account of sheaf cohomology would be highly desirable, not only
out of a philosophical desire to obtain a deeper understanding of the
foundations of sheaf cohomology, but also to: use the tools of sheaf cohomology
in the internal setting of toposes, thereby extending their applicability by
relativization by internalization; and to carry out \emph{integrated
developments} of algorithms for computing sheaf cohomology, where we would
extract algorithms together with termination and correctness proofs from a
hypothetical constructive account.

% XXX eff topos

\textbf{Acknowledgments.} We are grateful to Thorsten Altenkirch, Simon Henry
and Maria Emilia Maietti for insightful discussions and pointers to prior work,
to Jürgen Jost, Marc Nieper-Wißkirchen and Peter Schuster for valuable
guidance, and to Daniel Albert and Caterina Cozzi for their careful readings of
earlier drafts. Most of the work for this note was carried out at the University of
Augsburg and the Max Planck Institute for Mathematics in the Sciences in
Leipzig.

@@

There is an alternative definition of flabbiness, to be introduced below, which
is equivalent to the usual one in presence of Zorn's lemma and which
requires different commitments: For the alternative definition,
properties~(1),~(3b) and~(4a) can be verified purely intuitionistically.
There is a substitute for property~(3a) which can be verified purely
intuitionistically. We do not know whether property~(4b) can be established
purely intuitionistically, but we give a rudimentary analysis in
Section~\ref{sect:conclusion}.

@@

\begin{rem}Although we will see in Section~\ref{sect:in-eff} that there is some
relation between flabby sets and~$\neg\neg$-separated sets, neither notion
encompasses the other. The set~$\Omega$ is flabby, but might fail to
be~$\neg\neg$-separated; the set~$\ZZ$ is~$\neg\neg$-separated, even discrete,
but might fail to be flabby. This can abstractly be seen by adapting the proof of
Proposition~\ref{prop:taboo}. An explicit model in which~$\ZZ$ is not flabby
can be obtained by picking any topological space~$T$ such that~$H^1(T,
\underline{\ZZ}) \neq 0$. Then the constant sheaf~$\underline{\ZZ}$ is not
flabby and hence, by Proposition~\ref{prop:flabby-sheaves-objects} below, not a
flabby set from the internal point of view of~$\Sh(T)$.
\end{rem}

\begin{defn}\begin{enumerate}
\item A set~$I$ is \emph{injective} if and only if, for any injection~$i
: A \to B$, any map~$f : A \to I$ can be extended to a map~$B \to I$.
\item An~$R$-module~$I$ is \emph{injective} if and only if, for any linear
injection~$i : A \to B$ between~$R$-modules, any linear map~$f : A \to I$ can
be extended to a linear map~$B \to I$, as in the diagram below.
\end{enumerate}
\[ \xymatrix{
  A \ar@{^{(}->}[r]\ar[d] & B \ar@{-->}@/^/[ld] \\
  I
} \]
\end{defn}

@@

\begin{proof}\begin{enumerate}
\item One can check that, for instance, the full powerset~$\P(X)$ and the set of
subterminals~$\P_{\leq 1}(X)$ are each injective.
\item The forgetful functor from modules to sets possesses a left exact left
adjoint. More explicitly, if~$i : A \to B$ is an injective map between sets and
if~$f : A \to I$ is an arbitrary map, then the induced map~$R\langle A \rangle
\to R\langle B \rangle$ between free modules is also injective, the given
map~$f$ lifts to a linear map~$R\langle A \rangle \to I$, and an~$R$-linear
extension~$R\langle B \rangle \to I$ induces an extension~$B \to I$ of~$f$.
\item One verifies that any abelian group embeds into a divisible
abelian group. By Baer's criterion (which requires the axiom of choice), divisible abelian groups are injective.
The result for modules over arbitrary rings then follows purely
formally, since the functor~$A \mapsto \Hom(R,A)$ from abelian groups
to~$R$-modules has a left exact left adjoint with monic unit. \qedhere
\end{enumerate}\end{proof}

@@

Noticeably missing here is a statement as follows: ``Let~$0 \to M' \to M \to
M'' \to 0$ be a short exact sequence of modules. If~$M'$ and~$M$ are flabby, so
is~$M''$.'' Assuming Zorn's lemma in the metatheory, this statement is true in
every topos of sheaves over a locale, but we do not know whether it has an
intuitionistic proof and in fact we surmise that it has not.

@@

The assumption on~$f_*$ of Proposition~\ref{prop:pushforward-of-flabby-objects}
is for instance satisfied if~$f$ is a local geometric morphism.

@@

We originally set out to develop an intuitionistic account of Grothendieck's
sheaf cohomology. Čech methods can be carried out constructively, and
there are constructive accounts of special cases, resulting even in
efficient-in-practice algorithms~\cite{barakat-lh:homalg,barakat-lh:ext},
but it appears that there is not a general framework for sheaf
cohomology which works in an intuitionistic metatheory.

@@

However, it is an open question under which circumstances quotient inductive
types can be shown to exist. Zermelo--Fraenkel with choice certainly suffices,
while Zermelo--Fraenkel without choice does not~\cite[Section~9]{shulman-lumsdaine:hits},
hence IZF also does not.\footnote{With quotient inductive types, any infinitary
algebraic theory admits free algebras. However, it is consistent with
Zermelo--Fraenkel set theory that some such theories do not admit free
algebras~\cite{blass:free-algebras}.} The existence of quotient inductive types
seem to be, as the existence of enough injective modules, \emph{constructively
neutral}.

@@

\printbibliography
\enlargethispage{1em}
